---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# grantham

<!-- badges: start -->
<!-- badges: end -->

The goal of `{grantham}` is to provide a minimal set of routines to calculate
the Grantham distance [1].

The Grantham distance attempts to provide a proxy for the evolutionary distance
between two amino acids based on three key chemical properties: composition,
polarity and molecular volume. In turn, evolutionary distance is used as a proxy
for the impact of missense mutations. The higher the distance, the more
deleterious the substitution is expected to be.

## Installation

You can install the development version of `{grantham}` like so:

``` r
# install.packages("remotes")
remotes::install_github("maialab/grantham")
```

## Usage

Grantham distance between two amino acids:

```{r}
library(grantham)

grantham_distance(x = 'Ser', y = 'Phe')
```

The function `grantham_distance()` is vectorised with amino acids being matched element-wise to form pairs for comparison:

```{r}
grantham_distance(x = c('Ser', 'Arg'), y = c('Phe', 'Leu'))
```

The two vectors of amino acids must have compatible sizes in the sense of
[vec_recycle()](https://vctrs.r-lib.org/reference/vec_recycle.html) for element
recycling to be possible, i.e., either the two vectors have the same length, or
one of them is of length one, and it is recycled up to the length of the other.

```{r}
# `'Ser'` is recycled to match the length of the second vector, i.e. 3.
grantham_distance(x = 'Ser', y = c('Phe', 'Leu', 'Arg'))
```

Use the function `amino_acid_pairs()` to generate all 20 x 20 amino acid pairs:

```{r}
aa_pairs <- amino_acid_pairs()
aa_pairs
```

And now calculate all Grantham distances for all pairs `aa_pairs`:

```{r}
grantham_distance(x = aa_pairs$x, y = aa_pairs$y)
```

Because distances are symmetric, and pairs formed by the same amino acid are
trivially zero, you might want to exclude these pairs:

```{r}
# `keep_self = FALSE`: excludes pairs such as ("Ser", "Ser")
# `keep_reverses = FALSE`: excludes reversed pairs, e.g. ("Arg", "Ser") will be
# removed because ("Ser", "Arg") already exists.
aa_pairs <- amino_acid_pairs(keep_self = FALSE, keep_reverses = FALSE)

# These amino acid pairs are the 190 pairs shown in Table 2 of Grantham's
# original publication.
aa_pairs

# Grantham distance for the 190 unique amino acid pairs
grantham_distance(x = aa_pairs$x, y = aa_pairs$y)
```

The Grantham distance $d_{i,j}$ for two amino acids $i$ and $j$ is:

$$d_{i,j} = \rho (\alpha (c_i-c_j)^2+\beta (p_i-p_j)^2+ \gamma (v_i-v_j)^2)^{1/2}\ .$$

The distance is based on three chemical properties of amino acid side chains:

- composition ($c$)
- polarity ($p$)
- molecular volume ($v$)

We provide a data set with these properties:

```{r}
amino_acids_properties
```

If you want to calculate the Grantham distance from these property values you
may use the function `grantham_equation()`.


## Related software

Other sources we've found in the R ecosystem that also provide code for calculation of the Grantham distance:

- A GitHub Gist by Daniel E Cook provides the function `calculate_grantham()`, see [Fetch_Grantham.R](https://gist.github.com/danielecook/501f03650bca6a3db31ff3af2d413d2a).
- The `{midasHLA}` package includes the unexported function `distGrantham()` in [utils.R](https://github.com/Genentech/midasHLA/blob/ec29296f9bfd7c4fae9e2040592b618e5f2a99a1/R/utils.R).
- The `{HLAdivR}` package exports a data set with the Grantham distances in the format of a matrix, see [data.R]( https://github.com/rbentham/HLAdivR/blob/master/R/data.R).

## Code of Conduct
  
Please note that the `{grantham}` package is released with a [Contributor Code
of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.


## References

1. Grantham, R. _Amino acid difference formula to help explain protein evolution_. Science 185, 862--864
(1974). doi: [10.1126/science.185.4154.862](https://doi.org/10.1126/science.185.4154.862).
